---
title: "Stats140SL Project"
author: "Mrinalini Garg"
date: "11/27/2017"
output: html_document
---

```{r}
library(dplyr)
library(stringr)
library(rvest)
library( data.table)
library(corrplot)
library(lubridate)
library(ggplot2)
```


##Reading in all data sets  and exploratory data analysis 
```{r, eval=F}

cal<- read.csv("calendar.csv.gz") #Detailed Calendar Data for listings in Los Angeles

head(cal)
dim(cal)
names(cal)
str(cal)
cal$date<- ymd(cal$date)
summary(cal)

listing<- read.csv("listings.csv.gz") # Detailed listing data
head(listing$price)
dim(listing)
names(listing)
str(listing$price)
length(unique(listing$neighbourhood)) ## Only 160 neighbourhoods have data-- somehow differently named neighbourhoods 

listing$price<- str_replace( listing$price, "\\$", "")
listing$price<- as.numeric(str_replace( listing$price, ",", ""))

listing %>%  group_by(neighbourhood) %>%  summarize(revs= sum(number_of_reviews), avgrating= mean(review_scores_rating, na.rm=T), avg_price= mean(price, na.rm=T), count= n()) %>%  arrange(desc(revs)) # Now join with nbr and income table



listsum<- read.csv("listings.csv" )#Summary information and metrics for listings in Los Angeles 
dim(listsum)
names(listsum)
str(listsum)
length(unique(listsum$neighbourhood))
listsum %>%  group_by(neighbourhood) %>%  summarize(revs= sum(number_of_reviews)/n(), avg_price= mean(price, na.rm=T), count= n()) %>%  arrange(desc(revs))-> nbr_prc
summary(listsum[, c(6, 9, 10, 11, 12, 14,15, 16)])


# When did each listing get a review -- so dates for each listing 
review<- read.csv("reviews.csv")#Summary Review data and Listing ID (to facilitate time based analytics and visualisations linked to a listing).
str(review)
length(unique(review$listing_id))
dim(review)
review %>% group_by(listing_id) %>% summarise(revcount= n())


# Just a list of neighbourhoods
nbr<- read.csv("neighbourhoods.csv") #Neighbourhood list for geo filter. Sourced from city or open source GIS files
dim(nbr)
str(nbr)

```
 
 
## Scraping income data from la times 

```{r}
library(stringr)
library( rvest)
# Store web url
nbrincome <- read_html("http://maps.latimes.com/neighborhoods/income/median/neighborhood/list/")

#Scrape the website
nbr <- nbrincome %>% 
    html_nodes("#sortable_table a") %>%
    html_text() 
length(nbr)

inc<- nbrincome %>%  html_nodes("td:nth-child(3)") %>% html_text() 

nbr_inc<- as.data.frame( cbind( nbr, inc))
nbr_inc$inc %>% str_replace("\\$", "") %>% str_replace(",", "") %>%  as.numeric() -> nbr_inc$inc
head(nbr_inc)
str(nbr_inc)

```

# Data analysis

```{r}
setdiff( nbr_inc$nbr, listsum$neighbourhood)
#Data.tabe package: 
nbr_inc<- data.table( nbr_inc)
nbr_prc<- data.table( nbr_prc)

# joininh the 2 tables:
colnames(nbr_prc)[1]<- "nbr"
setkey(nbr_prc ,  nbr)
setkey(nbr_inc , nbr)

# perform the join using the merge function
Result <- merge(nbr_prc,nbr_inc, all.x=TRUE, by= "nbr")
head(Result)
str(Result)
Result<-Result[!(is.na( Result$inc)),]
summary(Result)

# Generating corrpolot 
M <- cor(Result[, c(2:4)])
corrplot.mixed(M, bg= "blue", upper= "square", )
M
View(Result)

plot( revs~ I(avg_price^2), data= Result[Result$avg_price<4000,])

md<- lm(revs~avg_price, data= Result)
plot(md)
summary(md)

hist(Result[Result$revs<10000]$revs)
hist(Result$revs)
sum(Result$revs==1)

md<- lm(revs~inc, data= Result)
plot(md)
summary(md)
plot(revs~inc, data= Result[Result$revs<1000])
sum(Result$revs==0)
sum(Result$count==0)
summary(lm( Result[Result$avg_price< 400]$inc~Result[Result$avg_price< 400]$avg_price))


md2<- lm(Result$inc~ log(Result$avg_price))
summary(md2)

hist(Result[Result$avg_price<500]$avg_price)

hist(Result$inc)



```


```{r}

plot(Result$avg_price ,Result$revs)
plot(Result$inc ,Result$revs)
plot(Result$count ,Result$revs)
plot(Result$avg_price ,Result$inc)


# Basic summary stats

# Linear relationships:
md<- lm(revs~(avg_price), data= Result)
plot(md)
summary(md)


## Behaviour accross different price bands and income ranges is very different and needs to be further segmented 

# Divding data by 3 income regions 

quantile( Result$inc, c(.33, .67, 1))

low<- Result[Result$inc<53312,]
name<- rep("low", nrow(low))
low<- cbind(low, name)
mid<- Result[Result$inc>53312 & Result$inc< 71998 ,]
name<- rep("mid", nrow(mid))
mid<- cbind(mid, name)
high<- Result[Result$inc>71999,]
name<- rep("high", nrow(high))
high<- cbind(high, name)

summary( low)
summary(high)
summary(mid)

dat<- rbind(low, mid, high)
dat$name<- as.factor(dat$name)
names(dat)[6]<- "Income"
dim(dat)
str(dat)

# And plotting 
ggplot(dat, aes(x=revs, colour= Income)) + geom_density(alpha= 0.2)
#Could do some non-parametric tests 
ggplot(dat, aes(x=avg_price, colour= Income)) + geom_density(alpha= 0.2)



plot(low$avg_price ,low$revs)
plot(low$inc ,low$revs)
plot(low$count ,low$revs)
plot(low$avg_price ,low$inc)


#Dividing data by price:

quantile( Result$avg_price, c(.33, .67, 1))

low<- Result[Result$avg_price<96.6,]
Range<- rep("low", nrow(low))
low<- cbind(low, Range)
mid<- Result[Result$avg_price>96.6 & Result$avg_price< 146.280 ,]
Range<- rep("mid", nrow(mid))
mid<- cbind(mid, Range)
high<- Result[Result$avg_price>146.3,]
Range<- rep("high", nrow(high))
high<- cbind(high, Range)

summary( low)
summary(high)
summary(mid)

dat1<- rbind(low, mid, high)
dat1$Range<- as.factor(dat1$Range)
dim(dat1)
str(dat1)

# And plotting 
ggplot(dat1, aes(x=revs, colour= Range)) + geom_density(alpha= 0.2)
#Could do some non-parametric tests 

ggplot(dat1, aes(x=inc, colour= Range)) + geom_density(alpha= 0.2)


ggplot(high, aes(x=inc)) + geom_density()

```


## Possible questions: 
How much are people willing to pay for each neighbourhood? or how is the price compared to the number of reviews and for each neighbourhood

are lower income neighbourhoods priced cheaper? 
How many reviews do they get? 
Are their reviews higher?
WHat kind of amenities do they offer 


Essentially do two linear regressions and assume the following- 
 1) NBR Income as x and number of reviews as Y ( so if correlated then good thing)
 2) Pricing as x and number of reviews as Y ( if correlated)
 3) Is there any correlation/ interaction between the 2 factores 
 4) How do they interact 
 OR 
 5) What amenities are people willing to lose out on for price changes- select essential amenities/ aggregate the different kinds
